Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-01-01 paper.Rnw
Added Yushkevich paper reviews

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
%  - remove 'accuracy' unless specifically talking relative to ground truth (and
%    even then, qualify)
%  - overlap/similarity --> agreement? choose one. 
%  - check that chart labels are sensible 
%  - in particular, check that Experiment 5 match Experiment 1
%  - fix figure 2 schematic language for the experiments (not 2a 2b). 
% 
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))
lm_eqn = function(m) {
  l <- list(a = format(coef(m)[1], digits = 2),
      b = format(abs(coef(m)[2]), digits = 2),
      pearson = format(sqrt(summary(m)$r.squared), digits = 3));
  if (coef(m)[2] >= 0)  {
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)~"="~pearson,l)
  } else {
    eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)~"="~pearson,l)    
  }

@@

Also of interest to this manuscript are methods that attempt to define
hippocampal subfields using standard T1- or T2-weighted data.  To the best of our
knowledge, there are only two automated segmentation algorithms that attempt
this problem. The first is included with the \freesurfer{} package
\citep{VanLeemput2009}. This work is limited as it omits the tail of the
hippocampus and the segmentation protocol has yet to be fully validated.
Nonetheless, it demonstrates that the applicability of hippocampal subfield
segmentation using data from 10 subjects. In the second method,
\citet{Yushkevich2009} hippocampal subfields were manually labelled on
high-resolution (either $0.2mm^3$ isotropic or $0.2mm \times 0.3mm \times 0.2mm$
resolution voxels) T2-weighted MR images acquired from five post-mortem medial
temporal lobe samples. Using nonlinear registration guided by shape-based models
of the subfield segmentations, and manually derived hippocampus masks of the
target images, the authors demonstrate accurate parcellation of hippocampal
subfields in clinical 3T T1-weighted MRI volumes. This method is not fully
automatic as it requires whole hippocampal parcellation of the target images.
%TODO: ref other Yush paper

In this paper we have describe a thorough validation of the \mb{} algorithm for
the automatic segmentation of the hippocampus and its subfields. First, we
address the feasibility of whole hippocampus segmentation with a limited number
of input atlases \citep{Chakravarty2013} by performing a multi-fold validation
experiment using a subset of the Alzheimer's Disease Neuroimaging Initiative
(ADNI) dataset manually segmented using the \prsnr{}-protocol, over a range of
atlas and template library sizes, registration and label fusion methods. We then
perform a leave-one-out validation to determine if hippocampal subfields can be
accurately identified, using the best parameters discovered in the first
experiment.  To ensure that we have not overfit our parameters to the aging or
neurodegenerative brain, we also apply \mb{} to a dataset of individuals
suffering from first episode psychosis.  Finally, we validate our algorithm
using all of the data available in the \adnidataset{} sample and compare our
segmentations to other popular segmentation algorithms.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
%  - remove 'accuracy' unless specifically talking relative to ground truth (and
%    even then, qualify)
%  - overlap/similarity --> agreement? choose one. 
%  - check that chart labels are sensible 
%  - in particular, check that Experiment 5 match Experiment 1
%  - fix figure 2 schematic language for the experiments (not 2a 2b). 
%  - ADNI-1 or ADNI1, standardize

<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))
lm_eqn = function(m) {
  l <- list(a = format(coef(m)[1], digits = 2),
      b = format(abs(coef(m)[2]), digits = 2),
      pearson = format(sqrt(summary(m)$r.squared), digits = 3));
  if (coef(m)[2] >= 0)  {
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)~"="~pearson,l)
  } else {
    eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)~"="~pearson,l)    
  }

@@

Also of interest to this manuscript are the methods that attempt to define
hippocampal subfields using standard T1- or T2-weighted data, of which there are
few.  \citet{VanLeemput2009} demonstrate that the applicability of hippocampal
subfield segmentation in T1-weighted images by using Markov random field priors
learned from 10 manual segmentations.  This work, available as part of the
\freesurfer{} package, is limited as the segmentation omits the tail of the
hippocampus and the protocol has yet to be fully validated.
\citet{Yushkevich2009} manually segment hippocampal subfields on high-resolution
(either $0.2mm^3$ isotropic or $0.2mm \times 0.3mm \times 0.2mm$ resolution
voxels) T2-weighted MR images acquired from five post-mortem medial temporal
lobe samples. Then, using nonlinear registration guided by shape-based models of
the subfield segmentations, and manually derived hippocampus masks of the target
images, the authors demonstrate accurate parcellation of hippocampal subfields
in clinical 3T T1-weighted MRI volumes.  Using \multiatlas with bias correction
techniques, \citet{Yushkevich2010} demonstrate a semi-automated method of
subfield segmentation on in vivo focal T2-weighted MR acquisitions of the
temporal lobe. Manual input is only needed to mark mark divisions between the
head, body and tail of the hippocampus on target images.

In this paper we describe a thorough validation of the \mb{} algorithm for the
fully automatic segmentation of the hippocampus and its subfields. First, we
address the feasibility of whole hippocampus segmentation with a limited number
of input atlases \citep{Chakravarty2013} by performing a multi-fold validation
experiment using a subset of the Alzheimer's Disease Neuroimaging Initiative
(ADNI) dataset manually segmented using the \prsnr{}-protocol, over a range of
atlas and template library sizes, registration and label fusion methods. We then
perform a leave-one-out validation to determine if hippocampal subfields can be
accurately identified, using the best parameters discovered in the first
experiment.  To ensure that we have not overfit our parameters to the aging or
neurodegenerative brain, we also apply \mb{} to a dataset of individuals
suffering from first episode psychosis.  Finally, we validate our algorithm
using all of the data available in the \adnidataset{} sample and compare our
segmentations to other popular segmentation algorithms.

@@

d = ddply(na.omit(wa.xval.casted), c('region'), function (df) {
  icc = ICC(df[4:5])
  data.frame(
    icc  = icc$results$ICC[3],
    f    = icc$results$F[3],
    lower= icc$results$"lower bound"[3],
    upper= icc$results$"upper bound"[3],
  )
})
